### 2024-03-23 thesis/appendix.tex
March 23rd 2024

>>>>

\section{Cross validation}\label{appendix:cross_validation}

\section{Quantile regressor extensive comparison}\label{appendix:quantile_regressor_extensive_comparison}
\subsection{Boston housing dataset}
The Boston housing dataset \href{https://www.kaggle.com/datasets/altavish/boston-housing-dataset}{https://www.kaggle.com/datasets/altavish/boston-housing-dataset} contains information about various attributes for suburbs in Boston.
There are 13 indipendent variables:
\begin{itemize}
\item CRIM per capita crime rate by town
\item ZN proportion of residential land zoned for lots over 25,000 sq.ft.
\item INDUS proportion of non-retail business acres per town.
\item CHAS Charles River dummy variable (1 if tract bounds river; 0 otherwise)
\item NOX nitric oxides concentration (parts per 10 million)
\item RM average number of rooms per dwelling
\item AGE proportion of owner-occupied units built prior to 1940
\item DIS weighted distances to five Boston employment centres
\item RAD index of accessibility to radial highways
\item TAX full-value property-tax rate per 10,000
\item PTRATIO pupil-teacher ratio by town
\item B $1000(Bk - 0.63)^2$ where Bk is the proportion of afroamericans by town
\item LSTAT lower status of the population
\end{itemize}
The dependent variable is MEDV, that is the median value of owner occupied homes in \$1000's

\begin{table}
\caption{Pinball loss Boston housing data}
\begin{tabular}{lllll}
\toprule
    & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
\midrule
0 & 13.785678 & 11.418540 & 10.587686 & 10.297572 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
    \caption{Pinball loss quantile-wise Boston data}
\begin{tabular}{lllll}
\toprule
    & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
\midrule
0.100000 & 0.729749 & 0.771714 & 0.588441 & 0.578898 \\
0.200000 & 1.122582 & 1.033442 & 0.932824 & 0.869145 \\
0.300000 & 1.479486 & 1.170642 & 1.153765 & 1.142783 \\
0.400000 & 1.712577 & 1.436263 & 1.352667 & 1.331955 \\
0.500000 & 1.911385 & 1.344361 & 1.408333 & 1.396300 \\
0.600000 & 1.989514 & 1.448885 & 1.464902 & 1.431705 \\
0.700000 & 1.938362 & 1.508741 & 1.427912 & 1.382772 \\
0.800000 & 1.658058 & 1.497901 & 1.275059 & 1.245288 \\
0.900000 & 1.243965 & 1.206591 & 0.983784 & 0.918725 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{MAE Boston data}    
\begin{tabular}{lllll}
\toprule
    & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
\midrule
0 & 3.826326 & 2.845989 & 2.965490 & 2.810494 \\
\bottomrule
\end{tabular}

\end{table}

\subsection{Abalone dataset}
The abalone data \href{https://archive.ics.uci.edu/dataset/1/abalone}{https://archive.ics.uci.edu/dataset/1/abalone} consist of measurements of abalone molluscs, the goal is predicting their age by building a model for estimating its number of rings; age is the number of rings plus 1.5
The data has 8 attributes:
\begin{itemize}
    \item Sex Categorical variable either male, female or infant
    \item Length
    \item Diameter
    \item Height
    \item Whole height
    \item Shucked height
    \item Viscera weight
    \item Shell weight
\end{itemize}

\begin{table}
    \caption{Pinball loss Abalone data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 5.613975 & 5.531938 & 5.212990 & 5.252491 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \caption{Pinball loss quantile-wise Abalone data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.277903 & 0.290531 & 0.274079 & 0.269287 \\
    0.200000 & 0.469361 & 0.488079 & 0.453110 & 0.457286 \\
    0.300000 & 0.621961 & 0.625633 & 0.580766 & 0.596791 \\
    0.400000 & 0.729757 & 0.715875 & 0.689904 & 0.691310 \\
    0.500000 & 0.794695 & 0.766185 & 0.735945 & 0.740834 \\
    0.600000 & 0.810691 & 0.785769 & 0.744928 & 0.746636 \\
    0.700000 & 0.769587 & 0.730318 & 0.700287 & 0.710392 \\
    0.800000 & 0.667776 & 0.656913 & 0.609378 & 0.608334 \\
    0.900000 & 0.472244 & 0.472635 & 0.424593 & 0.431621 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \caption{MAE Abalone data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 1.627627 & 1.574179 & 1.499522 & 1.498583 \\
    \bottomrule
    \end{tabular}
\end{table}

\subsection{Vehicle dataset}
This data contains info about used cars \href{https://www.kaggle.com/datasets/nehalbirla/vehicle-dataset-from-cardekho}{https://www.kaggle.com/datasets/nehalbirla/vehicle-dataset-from-cardekho}, the predictors are:
\begin{itemize}
    \item Year
    \item Present\_price ex showroom price
    \item Kms Driven
    \item Fuel type
    \item Seller type
    \item Transmission
    \item Owner number of previous owners
\end{itemize}
The dependent variable is the selling price.

\begin{table}
\caption{Pinball loss Vehicle data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 4.054449 & 2.289554 & 2.844410 & 2.204343 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \caption{Pinball loss quantile-wise Vehicle data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.254649 & 0.139849 & 0.170489 & 0.182490 \\
    0.200000 & 0.403772 & 0.236339 & 0.285875 & 0.223165 \\
    0.300000 & 0.548820 & 0.244086 & 0.357375 & 0.242963 \\
    0.400000 & 0.576918 & 0.263169 & 0.389305 & 0.262835 \\
    0.500000 & 0.554367 & 0.306123 & 0.410738 & 0.295878 \\
    0.600000 & 0.563046 & 0.335363 & 0.398125 & 0.306062 \\
    0.700000 & 0.516019 & 0.287490 & 0.326572 & 0.283716 \\
    0.800000 & 0.407742 & 0.261882 & 0.313698 & 0.235793 \\
    0.900000 & 0.229115 & 0.215253 & 0.192233 & 0.171440 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \caption{MAE Vehicle data}
    \begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0 & 1.117714 & 0.606971 & 0.752197 & 0.594292 \\
    \bottomrule
    \end{tabular}
\end{table}

### 2024-03-23 thesis/input_files/probabilistic.tex
March 23rd 2024

The dataset used is the Melbourne Daily Maximum Temperatures \cite{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days.

@@

- metrics pinball, picw plot, mae error on test data

>>>>

The dataset used is the Melbourne Daily Maximum Temperatures \cite{hdrcde}. It contains the daily maximum temperatures in Melbourne, Australia, from 1981-1990, excluding leap days, see figure \ref{fig:melbourne_temperature_data}.

@@

\begin{figure}
    \includegraphics[width=\textwidth]{images/melbourne_temperature.png}
    \caption{Melbourne temperatures dataset}
    \label{fig:melbourne_temperature_data}
\end{figure}
Hereafter, the results of the four presented methods on the Melbourne dataset are reported, see figure \ref{fig:melbourne_quantiles_comparison} for a visualisation. Hyperparameter tuning have been optimised through cross validation, see \ref{appendix:cross_validation}.
\begin{figure}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_linear_quantile_regression.png} 
      \caption{Linear quantile regression} 
      \label{fig:melbourne_linear_quantile_regression} 
      \vspace{4ex}
    \end{subfigure}%% 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_gradient_boosting_quantile_regression.png} 
      \caption{Gradient boosting quantile regression} 
      \label{fig:melbourne_gradient_boosting_quantile_regression} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melbourne_quantile_forest.png} 
      \caption{Quantile forest} 
      \label{fig:melbourne_quantile_forest} 
    \end{subfigure}%%
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[width=1.1\textwidth]{images/melborune_kernel_quantile_regression.png} 
      \caption{Kernel quantile regression} 
      \label{fig:melborune_kernel_quantile_regression} 
    \end{subfigure} 
    \caption{Quantile regressors}
    \label{fig:melbourne_quantiles_comparison} 
  \end{figure}

% tables
\begin{table}
\caption{Pinball loss Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    Pinball loss & 11.278895 & 10.317612 & 10.340842 & 10.031708 \\
    \bottomrule
    \end{tabular}
\end{table}

\begin{table}
    \caption{Pinball loss quantile-wise Melbourne data}
    \begin{tabular}{lllll}
    \toprule
    Quantiles & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    0.100000 & 0.710644 & 0.549232 & 0.562888 & 0.540235 \\
    0.200000 & 1.155014 & 0.938561 & 0.946712 & 0.903213 \\
    0.300000 & 1.417805 & 1.212671 & 1.222407 & 1.173803 \\
    0.400000 & 1.540108 & 1.399925 & 1.409293 & 1.367943 \\
    0.500000 & 1.574957 & 1.517281 & 1.484589 & 1.455849 \\
    0.600000 & 1.525114 & 1.498608 & 1.495474 & 1.447470 \\
    0.700000 & 1.397918 & 1.372183 & 1.362173 & 1.331503 \\
    0.800000 & 1.170140 & 1.115077 & 1.123195 & 1.096606 \\
    0.900000 & 0.787195 & 0.714075 & 0.734112 & 0.715085 \\
    \bottomrule
    \end{tabular}
\end{table}

As already pointed out, the quantile regression with q=0.5 corresponds to the standard regression problem, hence we can compare the proposed methods also in terms of the mean absolute error.

\begin{table}
\caption{MAE Melbourne data}
\begin{tabular}{lllll}
    \toprule
     & Linear qr & Gbm qr & Quantile forest & Kernel qr \\
    \midrule
    MAE & 3.253882 & 3.134805 & 3.095041 & 3.024336 \\
    \bottomrule
    \end{tabular}
\end{table}  
From these tables, we can see that kernel quantile regression outperformed the simple quantile regressor as well as the more complex models like quantile forest and gradient boosting quantile regression for the Melbourne temperatures dataset. Not only kernel quantile regression was the best in terms of total pinball loss but also the best in terms of each quantile pinball loss and mean absolute error.
Comparison has been carried out on further dataset, yielding similar conclusion as the one of above, to know more see \ref{appendix:quantile_regressor_extensive_comparison}.
